Black swans in ecological time series

### 2014-09-25 anderson-etal-blackswan-timeseries.tex
Blanket update to code and paper

\documentclass[10pt]{article}
\usepackage{geometry}
\geometry{letterpaper}
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsfonts}
%\usepackage{pdfpages}
%\usepackage{lscape}
%\usepackage{pdflscape}
%\usepackage{setspace}
\usepackage{booktabs}

@@

\mathchardef\mhyphen="2D
%\usepackage{url}
%\urlstyle{rm}

\textheight 20.5cm

@@

\title{The evidence for heavy tails in global\\population dynamics}
\author{Sean Anderson \ldots}
%\date{}

%\pagestyle{empty}
%\setcounter{secnumdepth}{-2}

\begin{document}

\input{values}

%\begin{spacing}{1.1}
%\doublespacing
%\onehalfspacing
\maketitle

\section*{Abstract}

\section*{Introduction}

Increasing realization about the importance of extreme events in the environment (REFs)

and of the importance of ecological surprises and black swans (REFs)

@@

\section*{Methods}

First, I tried fitting this as a Gompertz model and estimating the density dependence parameter. The Gompertz was by far the most favoured model in \citet{brook2006} with the GPDD.

\subsection*{Modelling framework}

The autoregressive Student-t Gompertz model is:
\begin{align*}
x_t &= \lambda + b x_{t-1} + \epsilon_t\\
\epsilon_t &\sim \mathrm{Student\mhyphen t}_\nu(\phi \epsilon_{t-1}, \sigma_\mathrm{proc}^2),
\end{align*}

\noindent where $x_t$ is the $\ln$ abundance at time $t$ and $\nu$ is the Student-t distribution degrees of freedom parameter. If $\nu$ is small (say 2--20) the distribution has heavy tails. As $\nu$ approaches infinity the distribution approaches the normal distribution. The parameter $b$ is the density dependence parameter. The model is density independent if $b = 1$ and maximally density dependent if $b = 0$. The parameter $\phi$ represent the first-order autoregressive correlation of the residuals $\epsilon$. $\sigma_\mathrm{proc}$ is the process error standard deviation. $\lambda$ represents the population growth rate per time step.

The mostly weakly-informative priors I'm using are:
\begin{align*}
b &\sim \mathrm{Uniform}(-1.2, 1.2)\\
\lambda &\sim \mathrm{Normal}(0, 20^2)\\
\nu &\sim \mathrm{Exponential}(0.01)\\
\phi &\sim \mathrm{Truncated\mhyphen Normal}(0, 1, \mathrm{min.} = -0.99, \mathrm{max.} = 0.99)\\
\sigma_\mathrm{proc} &\sim \mathrm{Half\mhyphen Cauchy} (5).
\end{align*}

Priors justification:

\begin{itemize}
\item $b$ is bounded just past stationary so we can detect if they are non-stationary will keeping the sampler from wandering off

\item $\lambda$ prior (variance = 400) is basically uninformative within the range of expected values for population growth... it allows even a X probability at a value of X

Using Stan 2.4.0 \citep{stan-manual2014}, and R 3.1.1 \citep{r2014} I'm starting with 4 chains and 2000  iterations with the first 1000 as warmup (i.e.\ 4000 total samples). If rhat is greater than 1.05 for any parameter or the minimum effective sample size is less than 200 for any parameter then I double both the total iterations and warmup and run again. These conditions were met by 8000 iterations (16000 total samples) in all cases and in almost all cases (X\%) within the initial 2000 iteration case (i.e.\ a total of 8000 samples).

Then, I again tried allowing for a specified level of observation error. I'm still running that now for just the mammals. From before, I believe this will reduce the probability of heavy tails a bit for many of them, but the vast majority that had heavy tails would still be considered heavy tailed. This does reduce substantially many of the bird heavy tails though.

(See the simulation results... I don't think this has the intended benefit and can probably be left out of the main results.)

When I incorporate observation error, the model is:
\begin{align*}
U_t &= \lambda + b U_{t-1} + \epsilon_t\\
x_t &\sim \mathrm{Normal}(U_t, \sigma_\mathrm{obs}^2)\\
\epsilon_t &\sim \mathrm{Student\mhyphen t}_\nu(\phi \epsilon_{t-1}, \sigma_\mathrm{proc}^2),
\end{align*}

\noindent
where $U$ represents the unobserved state vector, and $\sigma_\mathrm{obs}$ represents the standard deviation of observation error (on a log scale), which was set at 0.2.

\subsection*{Data selection}

Filtering rules for the GPDD:
We removed all populations with zeros (can't be fit with a Gompertz), all populations with uneven sampling intervals (end date minus start date for each step), and all populations with population estimates listed as ``harvest'' as well as those listed as from two fur trading sources (give details).
Then, we removed any populations with 5 or more identical values in a row since these suggest either recording error or extrapolation between two observations.
Then, we filled in all missing time steps with NAs and imputed single missing values with the geometric mean of the previous and following values (linear on the log scale that the Gompertz operates on). This leaves NAs that were recorded in successive time steps.
Then, we went through each population time series and found the longest window of abundance with at least 25 time steps. If there were any with multiple windows of identical length, we took the most recent window.
Finally, we removed main\_id 20531, which is a duplicate of 10139 (a heron population) and main\_id 10008, which had obvious errors in the data recording (a water vole population). We provide a supplemental figure of all the time series included in our analysis and indicate which values were imputed (\percImputedPops\% of populations had at least one point imputed and only \percImputedPoints\% of the total observations were imputed) (Fig.\ SX).

\subsection*{Simulation testing the model}

\section*{Results}

Main text figures:

\begin{enumerate}
\item illustrate $\nu$ and show simulated Gompertz time series with these levels of $\nu$ (include probability density at some extreme deviation in the caption or on figure for different $\nu$ values

\item a figure showing $\nu$ estimates within classes and  orders, ordered by decreasing $\nu$ estimate to illustrate percentage of populations, grouped by order in a phylogenetically related way include some silhouettes

\item 5--10 example time series with heavy tails from the GPDD; highlight the ``black swans'' with, say, red dots (perhaps also highlighting if they are up or down swings); pick some \textit{very} heavy tails and some moderately heavy tails. Show the $\nu$ estimate somewhere in each panel

\item something looking at hypotheses for what's associated with heavy tails: data collection method, mid date of time series (or when it started?), time series length, AR1, density dependence, lambda, body size for mammals, IUCN status? Most likely a grid of scatter plots coloured by taxonomic class.... or rows by taxonomic class and colours by order. (also ratio of generation time to collection window) (possibly, some of these will go in supplement with interesting ones in main paper).

Main text tables:

\begin{enumerate}
\item Table summarizing hypotheses for why we might see heavy tails
\item Pick some example populations (say the ones in the time series example plot) and show the nu estimate, reference, and a column digging into any known caveats or reasons
\end{enumerate}

Supplemental figures:

\section*{Discussion}

@@

\bibliography{/Users/seananderson/Dropbox/tex/jshort.bib,/Users/seananderson/Dropbox/tex/ref3.bib}

@@

\caption{TODO}\\

\toprule
Time series & Species & Black swan description & Reference \\
\midrule

\includegraphics[width=2cm]{sparks/6528} &
Shag (\textit{Phalacrocorax aristotelis}), Northumberland &
Shortage of nest sites reduced productivity; red-tide event in 1968 caused extreme mortality; no longer a nest shortage; population rapidly increased. &
\citep{potts1980}\\

%\includegraphics[width=2cm]{sparks/9675} &
%Carrot fly (\textit{Psila rosae}, Finland &
%TODO &
%\citep{markkula1965}\\

%\includegraphics[width=2cm]{sparks/10139} &
%Grey heron (\textit{Ardea cinerea}), UK &
%TODO &
%\citep{stafford1971}\\

\includegraphics[width=2cm]{sparks/1235} &
Wren (\textit{Troglodytes troglodytes}), UK &
TODO &
REF \\

%\includegraphics[width=2cm]{sparks/20580} &
%Chamois, \textit{Rupicapra rupicapra}, Switzerland &
%TODO &
%\citep{brook2006a}\\

\includegraphics[width=2cm]{sparks/5019} &
Barbary macaque, TODO &
TODO &
REF\\

@@

@@

\include{stat-table}
%\bibpunct{}{}{;}{a}{}{;}
%\begin{footnotesize}
%
%\renewcommand*{\arraystretch}{1.5}
%\begin{longtable}{>{\RaggedRight}p{3.5cm}>{\RaggedRight}p{8.2cm}>{\RaggedRight}p{2.2cm}}
%
%\caption{Description of black swans in GPDD}\\
%\toprule
%
%\midrule
%
%\bottomrule
%\label{tab:causes}
%\end{longtable}
%\end{footnotesize}
%\bibpunct{(}{)}{;}{a}{}{;}

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.6\textwidth]{t-nu.pdf}
\caption{A figure illustrating $\nu$. Maybe put some example (Gompertz) time series to the right with the same seed and different $\nu$ values?}
\label{default}
\end{center}
\end{figure}

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=\textwidth]{gomp-nu-ar1.pdf}
\caption{Estimate of $\nu$ from the Gompertz model (with estimated AR1 autocorrelation of the residuals but no observation error). Shown are median (points), quartiles (mid line segments) and 90\% quantiles (widest line segments) of the posterior.}
\label{fig:nu-coefs}
\end{center}
\end{figure}

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=\textwidth]{gomp-ar1-p20-dot-order.pdf}
\caption{Probability $\nu<20$, i.e.\ one way of asking what the probability is of heavy tails. Dots represent populations. Something like this could be made more pretty and sorted in some appropriate way\ldots maybe with little silhouettes for the animals. Looks pretty similar for $\nu < 10$, with the probabilities a bit lower. There is some vertical jitter added. Could do this and have another subpanel that breaks down the mammals by family, since there's the most interesting stuff happening there and there are lots of them... or just colour the dots by family within each panel.}
\label{fig:dots}
\end{center}
\end{figure}

%\clearpage
%
%\begin{figure}[htbp]
%\begin{center}
%\includegraphics[width=\textwidth]{gomp-ar1-p20-hist.pdf}
%\caption{Histogram of the same data as Figure~\ref{fig:dots} but lumped by class. Probably can't justify including this as well as the previous two figures unless there's some way we can make this more informative.}
%\label{default}
%\end{center}
%\end{figure}

%\clearpage

%\begin{figure}[htbp]
%\begin{center}
%\includegraphics[width=0.9\textwidth]{effect-of-ar1-on-nu.pdf}
%\caption{The effect of assuming no autocorrelation of residuals (vertical axis) vs.\ estimating autocorrelation of residuals (horizontal axis). I don't see a systematic effect. These are medians and interquartile ranges of the posterior. The colour shows $\phi$ --- the parameter estimating the lag-1 correlation of the residuals. The diagonal line illustrates 1:1. A few of these look strange (ones with $\phi$ near zero but far from the 1:1 line). It's possible the chains haven't converged sufficiently in these cases in one of the model versions. I will check what's going on and run more iterations in a final run if needed.}
%\label{default}
%\end{center}
%\end{figure}

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=\textwidth]{nu-covariates.pdf}
\caption{(Also see the next figure.) A number of hypotheses for what could be driving $\nu$ values. From top left to bottom right: $\phi$ (AR1 coefficient), $\log \sigma_\mathrm{proc}$ (log process error standard deviation), $N$ (time series length), $b$ (density dependence parameter in the Gompertz model; 1 is density independent and 0 is maximally density dependent), $\lambda$ (growth parameter from Gompertz model), $\log M$ (log adult body mass in grams; just for mammals and from the PanTHERIA database). All dots show median estimates and vertical line segments show interquartile range of posterior for the $\nu$ values. I don't generally see a whole lot going on here besides taxonomic differences but I'm open to some eye fuzzing and other ideas. If we want to, I was playing with a hierarchical beta regression with different levels for the taxonomic orders explaining the variation in p(nu $<$ 10) or these data\ldots but hypothesis generation with some visuals might be enough for this paper.}
\label{default}
\end{center}
\end{figure}

@@

%\begin{figure}[htbp]
%\begin{center}
%\includegraphics[width=\textwidth]{nu-covariates-p.pdf}
%\caption{Same as previous figure but the y axis is p(nu $<$ 20), which I think might be better to use\ldots, in reality this kind of coarse dichotomy I think is more realistic. If p(nu $<$ 20) is high (say above 0.5) then there's a good chance the tails are heavy. Otherwise, we just don't know.}
%\label{default}
%\end{center}
%\end{figure}

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=\textwidth]{ts-gpdd-heavy-eg-log10-no-fur.pdf}
\caption{Time series for populations with p(nu $<$ 10) $>$ 0.5. Ordered by increasing p(nu$<$10) (this value is shown pasted onto the front of each panel label). These are shown on a log10 y axis. There are a few funny time series we may want to look into and consider rules to remove. Colours show taxonomic class. This is for our understanding and a version of this could be a supplemental figure. But, we could take maybe 6 to 8 of these and use them as illustrations in the paper with the heavy tailed deviations highlighted. Could also probably dig into the literature on some of these and summarize in a table if we narrowed it down a bit. Lots of minks, beavers, and bears.}
\label{default}
\end{center}
\end{figure}

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=\textwidth]{t-dist-sampling-sim-prior-exp0point01.pdf}
\includegraphics[width=\textwidth]{t-dist-sampling-sim-sigma-prior-exp0point01.pdf}
\caption{TODO --- also maybe show the different exponential priors.}
\label{default}
\end{center}
\end{figure}

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=1.1\textwidth]{sim-gompertz.pdf}
\caption{\textbf{TODO Caption doesn't match new figure.} Results from a simulation test of the estimation model. Dashed horizontal lines are the true population value. True values were chosen to represent approximately the median values as estimated from the GPDD. ``Iteration'' refers to a stochastic draw from the true population distribution and a model fitting. The panels show true values of $\nu = $ 2.1, 5, 15, and 200 (very heavy, heavy, slightly heavy, and not heavy tailed). The panels from top to bottom show nu (red), process error standard deviation (blue), growth parameter (green), density dependence parameter (purple), and autocorrelation parameter (orange). The larger blocks of panels show four scenarios: (1) N = 100, (2) N = 50, (3) N = 50 with ignored observation standard deviation of 0.2, and (4) a model where the model also allows for observation standard deviation of 0.2
In general, this looks fairly good, with the caveat that if the tails are heavy, the model tends to overestimate process error standard deviation and underestimate just how heavy the tails are --- it's a bit biased towards calling heavy tails not as heavy as they are. Also, density independence gets consistently underestimated (purple). Observation error erodes the heavy-tail signal unless the tails are very heavy but doesn't tend to create false positives. Trying to allow for assumed observation error doesn't do much good besides fixing some very subtle bias in the process error estimate when the tails aren't heavy. We'd need a larger observation error to see this clearly. Line segments are 90\% quantiles of the posterior and the dots are the medians.}
\label{default}
\end{center}
\end{figure}

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=\textwidth]{check-sim-box.pdf}
\caption{Same data as previous figure but here represented as probability that nu is less than 10. Basically, this is looking at sensitivity/specificity. So, if you take something like a p = 0.5 that nu $<$ 10 as a cutoff, then you very rarely if ever get a false positive. However, the method doesn't \textit{always} detect heavy tails when they're there\ldots although in our data there are often tails that are much heavier than the nu = 2.5 used here.}
\label{default}
\end{center}
\end{figure}

% \begin{figure}[htbp]
%\begin{center}
%\includegraphics[width=0.6\textwidth]{effect-of-obs-on-nu.pdf}
%\caption{Effect of assuming that observation error is normally distributed (on a log scale) with standard deviation of 0.2.}
%\label{default}
%\end{center}
%\end{figure}
%

%\end{spacing}
\end{document}

>>>>

\documentclass[11pt]{article}
\usepackage{geometry}
\geometry{verbose,letterpaper,tmargin=2.54cm,bmargin=2.54cm,lmargin=2.54cm,rmargin=2.54cm}
\geometry{letterpaper}
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsfonts}
%\usepackage{pdfpages}
%\usepackage{lscape}
%\usepackage{pdflscape}
\usepackage{setspace}
\usepackage{booktabs}
\usepackage{lineno}

@@

\mathchardef\mhyphen="2D

\textheight 22.0cm

@@

\title{Black swans in ecological time series}
\author{
Sean C. Anderson$^{1\ast}$ \and
Trevor A. Branch$^2$ \and
Andrew B. Cooper$^3$ \and
Nicholas K. Dulvy$^1$
}
\date{}

% remove numbers in front of sections:
\makeatletter
\renewcommand\@seccntformat[1]{}
\makeatother

\begin{document}
\input{values} % R output

%\begin{spacing}{1.4}
%\doublespacing
\onehalfspacing

\maketitle
\thispagestyle{empty}

\noindent
\textit{Authors are listed alphabetically (ABCD!) and open to reordering.}

\noindent
\textsuperscript{1}Earth to Ocean Research Group, Department of Biological Sciences, Simon Fraser University, Burnaby BC, V5A 1S6, Canada

\noindent
\textsuperscript{2}School of Aquatic and Fishery Sciences, University of Washington, Box 355020, Seattle, WA 98195, USA

\noindent
\textsuperscript{3}School of Resource and Environmental Management, Simon Fraser University, Burnaby, BC, V5A 1S6, Canada

\noindent
\textsuperscript{*}Corresponding author: Sean C. Anderson; Earth to Ocean Research Group, Department of Biological Sciences, Simon Fraser University, Burnaby BC, V5A 1S6; Phone: 1-778-782-3989; E-mail: sean\_anderson@sfu.ca

\setcounter{page}{1}
\section{Abstract}

\section{Introduction}

Increasing realization about the importance of extreme events in the environment (REFs)
and of the importance of ecological surprises and black swans (REFs)

@@

\section{Methods}

\subsection{Modelling framework}

The Student-t Gompertz model is:
\begin{align*}
x_t &= \lambda + b x_{t-1} + \epsilon_t\\
%\epsilon_t &\sim \mathrm{Student\mhyphen t}_\nu(\phi \epsilon_{t-1}, \sigma_\mathrm{proc}^2),
\epsilon_t &\sim \mathrm{Student\mhyphen t}_\nu(0, \sigma_\mathrm{proc}),
\end{align*}

\noindent where $x_t$ is the $\ln$ abundance at time $t$. The model is density independent if $b = 1$ and maximally density dependent if $b = 0$. The parameter $\lambda$ represents the expected per capita growth rate at $\ln$ abundance = 0. The process noise is represented by a Student-t distribution that is centered at 0, has a scale parameter of $\sigma_\mathrm{proc}$, and has a degrees of freedom parameter of $\nu$. If $\nu$ is small ($\lesssim 10$) the distribution has heavy tails. As $\nu$ approaches infinity the distribution approaches the normal distribution (Fig~\ref{fig:didactic}).

We fit alternative models that allowed for autocorrelation of the residuals, allowed for observation error, and allowed the functional form of the population dynamics to be represented by a Ricker-logistic population model (Supporting Material).

\section{Results}

\section{Discussion}

@@

\bibliography{/Users/seananderson/Dropbox/tex/jshort,/Users/seananderson/Dropbox/tex/ref3}

@@

\caption{Example population dynamic black swans from the Global Population Dynamics Database and descriptions of their causes. Red and blue dots indicate downward and upward events that have a probability of occurring of 0.001\% or less if the population dynamics were explained by a Gompertz model with normally distributed process noise.}\\

\toprule
Time series (log scale) & Population & Black swan description & Reference \\
\midrule

\includegraphics[width=2cm]{sparks/6528} &
Shag,
\textit{Phalacrocorax aristotelis},
UK &
Shortage of nest sites reduced productivity; red-tide event in 1968 caused extreme mortality; no longer a nest shortage; population rapidly increased &
\citep{potts1980}\\

\includegraphics[width=2cm]{sparks/10007} &
Water vole,
\textit{Arvicola terrestris},
UK &
Short-term population cycles from predator interactions combined with longterm environmental cycle caused sharp downswing  &
\citep{saucy1994}\\

\includegraphics[width=2cm]{sparks/7115} &
Fur seal,
\textit{Arctocephalus pusillus},
South Africa &
Strong decreases in harvesting, loss of predators, and diamond mining regulations reducing human traffic caused sharp upswings  &
\citep{shaughnessy1982}\\

\includegraphics[width=2cm]{sparks/10113} &
Willow grouse,
\textit{Lagopus lagopus},
UK &
Parasite and predation effects interacted to cause low years  &
\citep{dobson1995}\\

\includegraphics[width=2cm]{sparks/10162} &
Red grouse,
\textit{Lagopus lagopus scoticus},
UK &
Good environmental conditions produced high numbers and vulnerable populations; bad conditions and overcrowding combined to create crashes  &
\citep{mackenzie1952}\\

\includegraphics[width=2cm]{sparks/1235} &
Wren,
\textit{Troglodytes troglodytes},
UK &
Severe winters where food was submerged in snow caused population crash &
\citep{newton1998} \\

\includegraphics[width=2cm]{sparks/20579} &
Grey heron,
\textit{Ardea cinerea},
UK &
Severe winters in 1929, 1940--1942, and 1962--1963; 1963 event so severe that recovery took three times as long as expected &
\citep{stafford1971} \\

%\includegraphics[width=2cm]{sparks/20580} &
%Chamois, \textit{Rupicapra rupicapra}, Switzerland &
% &
%\citep{brook2006a}\\

%\includegraphics[width=2cm]{sparks/5019} &
%Barbary macaque,  &
% &
%REF\\

%\includegraphics[width=2cm]{sparks/9675} &
%Carrot fly (\textit{Psila rosae}, Finland &
% &
%\citep{markkula1965}\\

%\includegraphics[width=2cm]{sparks/10139} &
%Grey heron (\textit{Ardea cinerea}), UK &
% &
%\citep{stafford1971}\\

@@

% reset citation style:
\bibpunct{(}{)}{;}{a}{}{;}

@@

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=\textwidth]{t-nu-eg.pdf}
\caption{An illustration of fitting population dynamics models that allow for heavy tails, represented by the Student-t degrees of freedom parameter $\nu$.
  (a, b) The probability density for t distributions with a scale parameter of 1 and different values of $\nu$.
  Small values of $\nu$ create heavy tails.
  As $\nu$ approaches infinity the distribution approaches the normal distribution.
  For example, the at $\nu = 2$, the probability of drawing a value less than -5 is 1.8\% whereas the probability of drawing such a value from a normal distribution is almost impossible: $2.9\cdot10^{-5}$\%.
  (c--e) Simulated population dynamics from a Gompertz model with process noise drawn from t distributions with different values of $\nu$.
  (f--h) Estimates of $\nu$ from models fit to the times series in panels c--e.
  Shown are the posterior samples (histograms), median and interquartile range of the posterior (IQR) (dots and line segments), and the exponential prior on $\nu$ (dashed lines).
  Colour shading behind panels f--h illustrates the region of heavy tails.
  Coloured dots in panels c and d represent points with less than a 1 in 1000 chance of occurring in a normal distribution.}
\label{fig:didactic}
\end{center}
\end{figure}

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=\textwidth]{gomp-nu-base.pdf}
\caption{
  Estimate of $\nu$ from the base Gompertz model (no autocorrelation or observation error).
  Shown are median (points), quartiles (mid line segments) and 90\% quantiles (widest line segments) of the posterior.}
\label{fig:nu-coefs}
\end{center}
\end{figure}

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=\textwidth]{gomp-base-p10-dot-order.pdf}
\caption{
  Probability $\nu<20$, i.e.\ one way of asking what the probability is of heavy tails.
  Dots represent populations.
  Something like this could be made more pretty and sorted in some appropriate way\ldots maybe with little silhouettes for the animals.
  Looks pretty similar for $\nu < 10$, with the probabilities a bit lower.
  There is some vertical jitter added. Could do this and have another subpanel that breaks down the mammals by family, since there's the most interesting stuff happening there and there are lots of them... or just colour the dots by family within each panel.}
\label{fig:dots}
\end{center}
\end{figure}

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=\textwidth]{nu-covariates-base.pdf}
\caption{
  (Also see the next figure.)
  A number of hypotheses for what could be driving $\nu$ values. From top left to bottom right: $\phi$ (AR1 coefficient), $\log \sigma_\mathrm{proc}$ (log process error standard deviation), $N$ (time series length), $b$ (density dependence parameter in the Gompertz model; 1 is density independent and 0 is maximally density dependent), $\lambda$ (growth parameter from Gompertz model), $\log M$ (log adult body mass in grams; just for mammals and from the PanTHERIA database).
  All dots show median estimates and vertical line segments show interquartile range of posterior for the $\nu$ values.
  I don't generally see a whole lot going on here besides taxonomic differences but I'm open to some eye fuzzing and other ideas.
  If we want to, I was playing with a hierarchical beta regression with different levels for the taxonomic orders explaining the variation in p(nu $<$ 10) or these data\ldots but hypothesis generation with some visuals might be enough for this paper.}
\label{fig:correlates}
\end{center}
\end{figure}

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=\textwidth]{brook-etal-correlates.pdf}
\caption{
  \citet{brook2006a} correlates plotted against nu estimates.
}
\label{fig:brook}
\end{center}
\end{figure}

\begin{centering}
\LARGE
Supporting material\\[1.5em]
\end{centering}

\section{Data selection}

TODO this paragraph is old:

Filtering rules for the GPDD:
We removed all populations with zeros NO LONGER TRUE (can't be fit with a Gompertz), all populations with uneven sampling intervals (end date minus start date for each step), and all populations with population estimates listed as ``harvest'' as well as those listed as from two fur trading sources (give details).
Then, we removed any populations with 5 or more identical values in a row since these suggest either recording error or extrapolation between two observations.
Then, we filled in all missing time steps with NAs and imputed single missing values with the geometric mean of the previous and following values (linear on the log scale that the Gompertz operates on). This leaves NAs that were recorded in successive time steps.
Then, we went through each population time series and found the longest window of abundance with at least 25 time steps. If there were any with multiple windows of identical length, we took the most recent window.
Finally, we removed main\_id 20531, which is a duplicate of 10139 (a heron population). We provide a supplemental figure of all the time series included in our analysis and indicate which values were imputed (\percImputedPops\% of populations had at least one point imputed and only \percImputedPoints\% of the total observations were imputed) (Fig.\ SX).

Go over zero rules.

\section{Details on heavy-tailed Gompertz probability model}

For the Gompertz, the mostly weakly-informative priors I'm using are:
\begin{align*}
b &\sim \mathrm{Uniform}(-1, 2)\\
\lambda &\sim \mathrm{Normal}(0, 10^2)\\
\nu &\sim \mathrm{Exponential}(0.01)\\
\phi &\sim \mathrm{Truncated\mhyphen Normal}(0, 1, \mathrm{min.} = -0.99, \mathrm{max.} = 0.99)\\
\sigma_\mathrm{proc} &\sim \mathrm{Half\mhyphen Cauchy} (5).
\end{align*}

See Figure~\ref{fig:priors}.

Priors justification: (TODO this is old)

\begin{itemize}
\item $b$ is bounded just past stationary so we can detect if they are non-stationary will keeping the sampler from wandering off too far

\item $\lambda$ prior (variance = 100) is basically uninformative within the range of expected values for population growth... it allows even a X probability at a value of X

Using Stan 2.4.0 \citep{stan-manual2014}, and R 3.1.1 \citep{r2014} I'm starting with 4 chains and 1000 iterations with the first 500 as warmup (i.e.\ 2000 total samples). If rhat is greater than 1.05 for any parameter or the minimum effective sample size is less than 200 for any parameter then I double both the total iterations and warmup and run again. These conditions were met by 8000 iterations (16000 total samples) in all cases and in almost all cases (X\%) within the initial 1000 iteration case (i.e.\ a total of 2000 samples).

\section{Simulation testing the model}

\section{Alternative population models}

\subsection{Ricker-logistic}

I also fitted a Ricker-logistic model:
\begin{align*}
x_t &= x_{t-1} + r_{\mathrm{max}}\left(1 - \frac{N_{t-1}}{K}\right) + b x_{t-1} + \epsilon_t\\
%\epsilon_t &\sim \mathrm{Student\mhyphen t}_\nu(\phi \epsilon_{t-1}, \sigma_\mathrm{proc}^2),
\epsilon_t &\sim \mathrm{Student\mhyphen t}_\nu(0, \sigma_\mathrm{proc}^2),
\end{align*}

Priors: For the Ricker-logistic $K$ uniform between 0 and twice the maximum observed abundance (\citet{clark2010} was less conservative at 0 and maximum observed). $r_\mathrm{max}$ uniform between 0 and 20 as in \citet{clark2010}. Same priors on $\nu$ and process standard deviation.

\subsection{Autocorrelated residuals}

\subsection{Assumed observation error}

When I incorporate observation error, the model is:
\begin{align*}
U_t &= \lambda + b U_{t-1} + \epsilon_t\\
x_t &\sim \mathrm{Normal}(U_t, \sigma_\mathrm{obs}^2)\\
\epsilon_t &\sim \mathrm{Student\mhyphen t}_\nu(0, \sigma_\mathrm{proc}^2),
\end{align*}

\noindent
where $U$ represents the unobserved state vector, and $\sigma_\mathrm{obs}$ represents the standard deviation of observation error (on a log scale), which was set at 0.3.

%\end{spacing}

\renewcommand{\thetable}{S\arabic{table}}
\setcounter{table}{0}

\include{stat-table}

\renewcommand{\thefigure}{S\arabic{figure}}
\setcounter{figure}{0}

% \begin{figure}[htbp]
% \begin{center}
% \includegraphics[width=\textwidth]{nu-covariates-p-base.pdf}
% \caption{Same as previous figure but the y axis is p(nu $<$ 20), which I think might be better to use\ldots, in reality this kind of coarse dichotomy I think is more realistic. If p(nu $<$ 20) is high (say above 0.5) then there's a good chance the tails are heavy. Otherwise, we just don't know.}
% \label{default}
% \end{center}
% \end{figure}

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=\textwidth]{ts-gpdd-heavy-eg-log10-no-fur.pdf}
\caption{Time series for populations with p($\nu < 10$) $>$ 0.5 using the base Gompertz population model. Panels are ordered by increasing p($\nu < 10$). Vertical axes are on a log10 scale. Colours indicate taxonomic classes. TODO add dots for upswings and downswings as in Table 1. The labels on each panel indicate p($\nu < 10$) $>$ 0.5, the common name for the species, and the GPDD ID number.}
\label{fig:heavy-ts}
\end{center}
\end{figure}

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.9\textwidth]{t-dist-sampling-sim-prior-exp0point01.pdf}
\includegraphics[width=0.9\textwidth]{t-dist-sampling-sim-sigma-prior-exp0point01.pdf}
\caption{
  Testing the ability to estimate $\nu$ (top panels) and the scale parameter of the process error (bottom panels) for a given number of samples (columns) drawn from a distribution with a given true $\nu$ value (rows). The red lines indicate the true population value. When a small number of samples are drawn there may not be samples sufficiently far into the tails to recapture the true $\nu$ value; however, heavy tails are still distinguished from normal tails in most cases, even with only 25 or 50 samples. TODO switch to nu = black swan, 3, 5, and normal.
}
\label{fig:sim-nu}
\end{center}
\end{figure}

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=1.1\textwidth]{sim-gompertz.pdf}
\caption{Simulation testing the Gompertz estimation model when the process error is chosen so that $\nu$ can be estimated close to the true value outside the full population model (``effective $\nu$'' within a CV of 0.5 of specified $\nu$).
  The simulation was run across population $\nu$ values (columns) and different scenarios (colours): (1) 100 time steps and no observation error, (2) 50 time steps and no observation error, (3) 50 time steps and observation error drawn from $\mathrm{Normal} (0, 0.3^2)$ but ignored, and (4) 50 time steps with observation error in which the quantity of observation error was assumed known.
  Dashed horizontal lines show the true population values;
  these true values were chosen to represent approximately the median values as estimated from the GPDD.
  Individual dots and lines represent a stochastic draws from the true population distribution and a model fitting.
  The panels show true values of $\nu$ =  3, 5, normal (very heavy, heavy, and not heavy tailed).
  The panels from top to bottom show $1/\nu$, process noise scale parameter $\sigma_\mathrm{proc}$, growth rate parameter $\lambda$, and the density dependence parameter $b$.
  Dots and line segments are medians and 80\% credible intervals of the posterior.}
\label{fig:sim-gompertz}
\end{center}
\end{figure}

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=\textwidth]{check-sim-box.pdf}
\caption{The same simulations results as the previous figure but here represented as probability that $\nu < 10$ (i.e.\ the probability of heavy tails). Within each scenario the dots represent stochastic draws from the true population distributions.}
\label{fig:sim-prob}
\end{center}
\end{figure}

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.9\textwidth]{gomp-vs-logistic-nu.pdf}
\caption{Gompertz vs.\ Ricker logistic $\nu$ (nu) estimates. Line segments show 50\% credible intervals; dots show median estimates. The vast majority of heavy-tailed estimates are similar for the two population models. In a few instances one model estimates heavy tails where the other does not. This happens more often for the Ricker-logistic than the Gompertz model.}
\label{fig:gomp-vs-ricker}
\end{center}
\end{figure}

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.8\textwidth]{effect-of-ar1-on-nu.pdf}
\caption{The effect of assuming no autocorrelation of residuals (vertical axis) vs.\ estimating autocorrelation of residuals (horizontal axis). Shown are medians and 50\% credible intervals of the posterior. The colours indicate $\phi$ --- the parameter estimating the lag-1 correlation of the residuals. The diagonal line illustrates a one-to-one relationship. There are a few populations in which the heavy-tail estimate changes when autocorrelation is incorporated, but a systematic difference.}
\label{fig:ar1}
\end{center}
\end{figure}

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.9\textwidth]{priors-gomp-base.pdf} \caption{Probability
  density of the Bayesian priors for the base Gompertz model. Per capita growth
  rate at log(abundance) = 0: $\lambda \sim \mathrm{Normal}(0, 10^2)$; scale
  parameter of t-distribution process error: $\sigma_\mathrm{proc} \sim \mathrm{Half\mhyphen
    Cauchy} (5)$; t-distribution degrees of freedom parameter: $\nu \sim
  \mathrm{Exponential}(0.01)$. Not shown is $b$, the density dependence
  parameter: $b \sim \mathrm{Uniform}(-1, 2)$.}

\label{fig:priors}
\end{center}
\end{figure}

\end{document}

1 - there is evidence of heavy tails/black swans... in ecological time series... but they are rare

- taxonomically widespread
- we don't find compelling... intrinsic obvious biological reason
- commonly extrinsic extreme events and interactions - climate and severe winters, cycles of natural enemies (parasites + predators),
- as climate gets more variable, extremes more frequent and greater magnitude (and skewed), might expect more of these
- and as our time series get longer and censusing gets more accurate... more ability to detect these
- calls for policy that is be robust to the possibility of black swans
- lead with ecological application

grouse are hunted - shags have erupted
